---
layout: page
title: "Life Expectancy Dashboard"
format: 
  dashboard:
    nav-buttons: [github]
    github: https://github.com/your-repo/study2-dashboard
theme: [sandstone]
params:
  dataset: "Study 2 Data"
editor_options: 
  chunk_output_type: console
---


```{r}
#| label: load-packages
#| message: false

library(tidyverse)
library(scales)
library(DT)
library(gt)
theme_set(theme_minimal(base_size = 16, base_family = "sans"))
```


```{r}
#| label: load-data

final_data <- readRDS("/Users/taravatkhodaii/Desktop/Project B new data/copy study 2 the final/study2_data_cleaned.rds")
```


```{r}
#| label: prep-data

final_data <- final_data |>
  mutate(
    severe_housing_cat = factor(
      severe_housing_cat,
      levels = c("Low", "Moderate", "High", "Very High"),
      labels = c("Low", "Moderate", "High", "Very High") # Adjust labels if needed
    )
  )

```


# Overview {.sidebar}

```{r}
#| label: sidebar-content

# Sidebar content setup
sidebar_content <- tibble(
  "Key Metrics" = c("Dataset", "Outcome Variable", "Key Predictor"),
  "Details" = c(
    "Study 2 Data",
    "Life Expectancy",
    "Severe Housing Problems"
  )
)

# Render as a table
library(knitr)
kable(
  sidebar_content,
  col.names = c("", ""),
  caption = "Key Details of Study 2"
)
```

# Key Statistics
```{r}
#| label: summary-stats
#| echo: false

# Summary table of dataset
summary_table <- final_data |>
  summarize(
    `Total Observations` = n(),
    `Mean Life Expectancy` = mean(life_expectancy, na.rm = TRUE),
    `Median Life Expectancy` = median(life_expectancy, na.rm = TRUE),
    `Min Life Expectancy` = min(life_expectancy, na.rm = TRUE),
    `Max Life Expectancy` = max(life_expectancy, na.rm = TRUE)
  )

gt(summary_table)

```

## Life Expectancy Distribution
```{r}
#| label: life-expectancy-dist
#| echo: false

# Histogram of life expectancy
ggplot(final_data, aes(x = life_expectancy)) +
  geom_histogram(binwidth = 1, fill = "blue", color = "orange", alpha = 0.7) +
  labs(
    title = "Distribution of Life Expectancy",
    x = "Life Expectancy (years)",
    y = "Frequency"
  ) +
  theme_minimal()

```


# Model Results
```{r}
#| label: model-results

# Fit models
full_model <- lm(life_expectancy ~ severe_housing_cat + physical_inactivity + 
                 income_inequality + air_pollution + social_associations + 
                 preventable_hospital_stays, data = final_data)

subset_model <- lm(life_expectancy ~ severe_housing_cat + physical_inactivity + 
                   income_inequality, data = final_data)

# Model summaries
full_model_summary <- summary(full_model)
subset_model_summary <- summary(subset_model)

# Display Adjusted R-squared values
performance_metrics <- tibble(
  Model = c("Full Model", "Subset Model"),
  `Adjusted R²` = c(full_model_summary$adj.r.squared, subset_model_summary$adj.r.squared)
)

gt(performance_metrics)

```

# Model Coefficients
```{r}
#| label: coef-plot
#| echo: false

# Extract coefficients for full model
coef_full <- broom::tidy(full_model)

# Coefficient plot
ggplot(coef_full, aes(x = term, y = estimate)) +
  geom_col(fill = "steelblue") +
  coord_flip() +
  labs(
    title = "Full Model Coefficients",
    x = "Predictor",
    y = "Estimate"
  ) +
  theme_minimal()

```



# Data Table 
```{r}
#| label: data-table
#| echo: false

datatable(
  final_data,
  options = list(pageLength = 10),
  rownames = FALSE
)

```

# Conclusion
---

## Key Findings

- **Physical Inactivity** was the strongest predictor of life expectancy, showing the most significant negative impact.
- **Income Inequality** and **Severe Housing Problems** also had meaningful impacts on life expectancy.

## Model Comparison

- The **Full Model** slightly outperformed the **Subset Model**:
  - Adjusted \(R^2\) for the Full Model: `r round(summary(full_model)$adj.r.squared, 3)`
  - Adjusted \(R^2\) for the Subset Model: `r round(summary(subset_model)$adj.r.squared, 3)`

- The Subset Model, being simpler, provides comparable interpretability while requiring fewer predictors.

## Implications

- Public health interventions targeting **physical inactivity** could have the most significant impact on improving life expectancy.
- Addressing **income inequality** and **housing problems** could further enhance societal health outcomes.


```{css echo=FALSE}
.embed-container {
    position: relative;
    padding-bottom: 129%;
    height: 0;
    overflow: hidden;
    max-width: 100%;
}
.embed-container iframe,
.embed-container object,
.embed-container embed {
    position: absolute;
    top: 0;
    left: 0;
    width: 100%;
    height: 100%;
}
```

```{=html}
<p class="text-center">
  <a class="btn btn-primary btn-lg cv-download" href="`r rmarkdown::metadata$cv$pdf`" target="_blank">
    <i class="fa-solid fa-file-arrow-down"></i>&ensp;Download current CV
  </a>
</p>
<div class="embed-container">
  <iframe src="`r rmarkdown::metadata$cv$pdf`#toolbar=0" style="border: 0.5px;width: 100%;"></iframe>
</div>
```





